Processing Southern Fish Data
========================================================


### Setup

Cleaning before entry into R:
* Replaced `,` with blank to read csv properly
* Replaced `'` and `"` with blank to avoid problems reading in text columns
* Replaced `ND` (No Data) with `NA` for missing data
* Changed `Species` to `Code` in the fish data headers to be able to merge with the species data (could do in R after import)

```{r setup, results = 'hide', message = FALSE, warning = FALSE}
library(data.table)
library(dplyr)
library(ggplot2)

setwd('/Users/Dan/Documents/Research/Stream_Climate_Change/Brook_Trout_Data/Southern/')

fish <- fread("Southern_Fish.csv", header = TRUE, sep = ",")
sites <- fread("Southern_Site_Info.csv", header = TRUE, sep = ",")
spp <- fread("Southern_Spp_Codes.csv", header = TRUE, sep = ",")
```

Add primary key for resorting back to original data

```{r primary key, results = 'hide'}
summary(fish)
fish$fish.key <- 1:length(fish$Pass)
sites$site.key <- 1:length(sites$SiteID)
spp$spp.key <- 1:length(spp$Code)
```

### Merge data sets, organize dates, and export

```{r}
data <- left_join(fish, sites, by = 'SiteID')
summary(data)
sites[which(is.na(sites$LonDD)), ]

data <- left_join(data, spp, by = 'Code')
summary(data)

data$date <- as.Date(data$Date, format = "%m/%d/%Y")
data$month <- as.numeric(format(data$date, format = "%m")) 
data$doy <- strptime(data$Date, "%m/%d/%Y")$yday + 1

head(data)

write.table(data, "All_Field_Data.csv", row.names = FALSE, sep = ",")

fish$date <- as.Date(fish$Date, format = '%m/%d/%Y')
trip <- summarise(group_by(fish, TripID),
                   Date = min(date),
                  Year = mean(Year))

trip$SiteID <- gsub("(.*)\\_.*", "\\1",  x = trip$TripID)
```

11 sites have not lat and lon recorded which results in 1272 records with missing location data out of 252,951 records (0.5%)

### Add stages

Plot histograms of all species lengths to get estimates of cutoffs for stage (hopefully doesn't have to be done by month)

```{r results = 'hide'}
# Plot to find cutoffs
for(i in 1:length(unique(fish$Code))) {
  g <- ggplot(filter(fish, Code == unique(fish$Code)[i]), aes(Length))
  print(g + geom_histogram(binwidth = 10) + ggtitle(unique(fish$Code)[i]) + scale_x_continuous(breaks = seq(0, max(fish[which(fish$Code == unique(fish$Code)[i]), ]$Length, na.rm = T), 20)))
  }

i = 13
g <- ggplot(filter(fish, Code == unique(fish$Code)[i] & Length < 200), aes(Length))
print(g + geom_histogram(binwidth = 5) + ggtitle(unique(fish$Code)[i]) + scale_x_continuous(breaks = seq(0, 200, 10)))

i = 14
g <- ggplot(filter(fish, Code == unique(fish$Code)[i]), aes(Length))
print(g + geom_histogram(binwidth = 10) + ggtitle(unique(fish$Code)[i]) + scale_x_continuous(breaks = seq(0, 200, 10)))
```

Potential Outlier Lengths:
* SAFO > 300
* SATR > ~350?
* ONMY > ~320?
* RHCA > 200
* COBA > 150

YOY cutoffs:
* SAFO: 90 mm
* LEGI: Insufficient Data
* AMRU: Unclear
* MIDO: 70 mm (not clear)
* AMNA: Insufficient Data
* SATR: 100 mm
* LEAU: Insufficient Data
* LEMA: Insufficient Data
* HXTR: Insufficient Data
* ONMY: 90 mm
* AMNE: Insufficient Data
* MISA: Insufficient Data
* RHCA: 20 mm
* RHAT: 10 mm
* COBA: 40 mm (maybe 45 mm)
* CAAN: Insufficient Data
* SEAT: Insufficient Data

```{r}
# Add stage
data$stage <- NA
data[which(data$Length < 90 & data$Code == 'SAFO'), ]$stage <- 0 # YOY
data$stage[which(data$Length >= 90 & data$Code == 'SAFO')] <- 1 # Adult

data[which(data$Length < 70 & data$Code == 'MIDO'), ]$stage <- 0 # YOY
data$stage[which(data$Length >= 70 & data$Code == 'MIDO')] <- 1 # Adult

data$stage[which(data$Length < 100 & data$Code == 'SATR')] <- 0 # YOY
data$stage[which(data$Length >= 100 & data$Code == 'SATR')] <- 1 # Adult

data$stage[which(data$Length < 90 & data$Code == 'ONMY')] <- 0 # YOY
data$stage[which(data$Length >= 90 & data$Code == 'ONMY')] <- 1 # Adult

data$stage[which(data$Length < 20 & data$Code == 'RHCA')] <- 0 # YOY
data$stage[which(data$Length >= 20 & data$Code == 'RHCA')] <- 1 # Adult

data$stage[which(data$Length < 10 & data$Code == 'RHAT')] <- 0 # YOY
data$stage[which(data$Length >= 10 & data$Code == 'RHAT')] <- 1 # Adult

data$stage[which(data$Length < 40 & data$Code == 'COBA')] <- 0 # YOY
data$stage[which(data$Length >= 40 & data$Code == 'COBA')] <- 1 # Adult
```

Convert to count data (problem of how to fill in zeros) - maybe should do this for all species before subsetting to bkt. Would have to assign stage wherever possible first

```{r}
# Remove any rows where pass = NA
data <- data[which(data$Pass != "NA"), ]
summary(data)

count <- summarise(group_by(data, TripID, Pass, Code), 
                   count = n(),
                   length.mean = mean(Length),
                   mass.mean = mean(Weight))

count.stage <- summarise(group_by(data, TripID, Pass, Code, stage), 
                   count = n(),
                   length.mean = mean(Length),
                   mass.mean = mean(Weight))
```

Expand data to include zeros for all species per trip and pass. Then merge back with trip and site information.

```{r}
# Expand to trip-species-pass
mergedData <- merge(
    expand.grid(TripID = unique(data$TripID), Code = unique(data$Code), Pass = unique(data$Pass)),
    y = count, by = c("TripID", "Code", "Pass"), all.x = T)
head(mergedData, 100)
summary(mergedData)

mergedData[is.na(mergedData$count), ]$count <- 0

# remerge/join the environmental variables
mergedData$TripID <- as.character(mergedData$TripID)

all.count <- left_join(mergedData, trip, by = 'TripID') # Add observational info (date)

all.count <- left_join(all.count, sites, by = 'SiteID')
all.count$Code <- as.character(all.count$Code)
all.count <- left_join(all.count, spp, by = 'Code')
str(all.count)
summary(all.count)
head(all.count, 25)

# Remove extra columns and rename remaining columns**********************

data$date <- as.Date(data$Date, format = "%m/%d/%Y")
data$month <- as.numeric(format(data$date, format = "%m")) 
data$doy <- strptime(data$Date, "%m/%d/%Y")$yday + 1

head(data)

# Need to do this for stage-based data too (then separate adults)
count.stage$TripID <- as.factor(count.stage$TripID)

mergedData <- merge(
    expand.grid(TripID = unique(data$TripID), Code = unique(data$Code), Pass = unique(data$Pass)),
    y = count.stage, by = c("TripID", "Code", "Pass"), all.x = T)
head(mergedData, 100)
summary(mergedData)

mergedData[is.na(mergedData$count), ]$count <- 0

# Need to remerge/join the environmental variables

# could probably do this better with SQL query coding
count.stage <- left_join(mergedData, data, by = 'TripID') # Add site & observational info (date)
count.stage$Code <- count.stage$Code.x
count.stage <- left_join(count.stage, spp, by = 'Code')
str(count.stage)
summary(data)
summary(count.stage)
head(count.stage, 25)
sites[which(is.na(sites$LonDD)), ]

summary(count.stage)
count.stage$Pass.x == count.stage$Pass.y

# Keep only useful columns, reorder, and rename
vars <- c("State", "SiteID", "LatDD", "LonDD", "TripID", "Pass.x", )
count.stage <

count.adult <- filter(count.stage, filter = stage == 1)

summary(count)
head(count, 50)
```

### Subset brook trout data and clean
 Consider keeping brown trout to use as a covariate

```{r}
summary(filter(fish, Species == 'SAFO')) # check lengths around min and max

ggplot(filter(fish, Species == 'SAFO' & Length < 250), aes(Length), ) + 
  geom_histogram(binwidth = 10) + 
  scale_x_continuous(breaks = seq(0, 250, 20))

safo <- filter(fish, Species == 'SAFO')

summarise(filter(group_by(data, Pass), Code == 'SAFO'), mean(Length, na.rm = TRUE)) # catch biggest SAFO first 
```
